### 2013-05-19 draft.tex
fixed the rest of the constraining temps section, havent read thru yet

% comparisons support hypothesis that low temp most pristine
% array shows rock-buffered diagenesis
% most pristine sample only one replicate
% only upper estimates because all samples could have a little contamination
% Silurian warm and similar d18O to today
% resolve veizer controversy
% surprising that d18O has not changed

@@

To estimate the temperature and \deltao O of Silurian oceans, we first need to confirm the hypothesis that the lowest-temperature samples are the most pristine and therefore record the most reliable record of Silurian ocean conditions. This hypothesis can be tested through constraints on the diagenetic regime to which the samples were subjected. As outlined above, different diagenetic regimes have characteristic effects on the relationship between temperature and the \deltao O of interstitial water. When the clumped isotope temperatures of the samples in this study are plotted versus the \deltao O of water calculated for each sample, there is a clear positive relationship (Figure \ref{ellipses}). This relationship suggests that interstitial water grew increasingly enriched in $^{18}$O as the temperature increased, which is diagnostic of rock-buffered diagenesis. Importantly, during rock-buffered diagenesis, more altered samples have higher temperatures, which is consistent with our hypothesis that the lowest temperature samples are the most pristine.

Solid-state diffusion is commonly considered as a possible mechanism for the increase of clumped isotope temperatures during diagenesis, but we rule it out as a major contributor to our results. During solid-state diffusion, isotopes of carbon and oxygen diffuse through the calcite mineral lattice, which decreases the degree of clumping and increases the interpreted temperature of the calcite. This process does not affect the bulk isotopic composition of the calcite, and it occurs without dissolution and reprecipitation, which means it is also not expected to change the calcite fabric or trace metal content \citep{Eiler2011}. However, solid-state diffusion is expected to affect all calcite approximately equally, which means that this process cannot explain the variation in temperature among the samples. Additionally, solid-state diffusion is thought to occur very slowly at temperatures below 100\degrees C (D. Stolper personal communication 2013), which is the maximum temperature experienced by Gotland strata, as constrained by the conodont alteration index \citep{Jeppsson1983,Wenzel2000}. Solid-state diffusion is not consistent with our observations, leaving rock-buffered diagenesis as the most likely characterization of the diagenetic regime.

Comparisons of the clumped isotope temperature of different calcite textures provide additional support for the hypothesis that the samples underwent rock-buffered diagenesis. We measured the clumped isotope composition of the micrite adjecent to four brachiopods, and in all four cases, the temperature of the micrite was resolvably higher than the temperature of the brachiopod (Figure \ref{comparisons}). Micrite is expected to be less pristine than the brachiopod samples because micrite is fine-grained, not protected by a layer of brachiopod shell, and was not picked under a microscope prior to analysis. Because the less pristine material, micrite, was associated with higher temperatures, these comparisons are consistent with the hypothesis that the lowest temperature samples are the most pristine.

@@

Our results, along with those of earlier clumped isotope studies \citep{Came2007,Finnegan2011,Dennis2013}, provide consistent explanations for the low \deltao O values of early Paleozoic carbonates. Carbonate clumped isotopes constrain the contribution of a change in seawater \deltao O to only a few permil of the observed trend, ruling out a value of $-8\permil$ \citep{Jaffres2007,Veizer1999} for the \deltao O of early Paleozoic oceans \citep{Grossman2012}. Our results also serve as a reminder that many of the fossils preserved in the geologic record lived in tropical oceans. The \deltao O of carbonate in equilibrium with tropical ocean water should not be compared to the \deltao O of carbonate in equilibrium with water of average modern sea surface temperature, because tropical oceans are much warmer than the average sea surface temperature. Clumped isotopes additionally demonstrate that extremely hot ($\sim 70$\degrees C) ocean temperatures cannot be the cause of the depleted \deltao O values of early Paleozoic carbonates. Finally, clumped isotopes, as a highly sensitive indicator of diagenetic alteration, demonstrate that even among samples with original calcite fabric and pristine trace metal signatures, there is a large range in preservation quality. This means that the average fossil calcite in almost any sample suite is likely altered from its primary composition, and that diagenetic alteration is a key contributor to the low \deltao O values of early Paleozoic carbonates.

Given the large potential for change in the \deltao O of seawater through exchange of oxygen isotopes with crustal material, it is surprising that four separate clumped isotope studies, each from a different time in the Phanerozoic, have all implied similar \deltao O values of seawater.  Several studies have suggested that high-temperature alteration of the seafloor acts as a buffer for the \deltao O of seawater \citep{Muehlenbachs1976,Gregory1981,Gregory1991,Muehlenbachs1998}, but a more recent model implied that this buffer may not be strong enough to overcome slow changes in the \deltao O of seawater \citep{Jaffres2007}. The controls on the magnitudes of seafloor and continental weathering do not form a clear negative feedback loop, so we would not expect the \deltao O of seawater to be broadly constant through time, contrary to the evidence of several clumped isotope studies.

>>>>

@@

To estimate the temperature and \deltao O of Silurian oceans, we first need to confirm the hypothesis that the lowest-temperature samples are the most pristine and therefore preserve the most reliable record of Silurian ocean conditions. This hypothesis can be tested through constraints on the diagenetic regime to which the samples were subjected. As outlined above, different diagenetic regimes have characteristic effects on the relationship between temperature and the \deltao O of interstitial water. When the clumped isotope temperatures of the samples in this study are plotted versus the \deltao O of water calculated for each sample, there is a clear positive relationship (Figure \ref{ellipses}). This relationship suggests that interstitial water grew increasingly enriched in $^{18}$O as the temperature increased, which is diagnostic of rock-buffered diagenesis. Importantly, during rock-buffered diagenesis, more altered samples have higher temperatures, which is consistent with our hypothesis that the lowest temperature samples are the most pristine.

Solid-state diffusion is commonly considered as a possible mechanism for the increase of clumped isotope temperatures during diagenesis, but we rule it out as a major contributor to our results. During solid-state diffusion, isotopes of carbon and oxygen diffuse through the calcite mineral lattice, which decreases the degree of clumping and increases the interpreted temperature of the calcite. This process does not affect the bulk isotopic composition of the calcite, and it occurs without dissolution and reprecipitation, which means it is also not expected to change the calcite fabric or trace metal content \citep{Eiler2011}. However, solid-state diffusion is expected to affect all calcite approximately equally, which means that this process cannot explain the observed variability in temperature among the samples. Additionally, solid-state diffusion is thought to occur very slowly at temperatures below 100\degrees C (D. Stolper personal communication 2013), which is the maximum temperature experienced by Gotland strata, as constrained by the conodont alteration index \citep{Jeppsson1983,Wenzel2000}. Solid-state diffusion is not consistent with our observations, leaving rock-buffered diagenesis as the most likely characterization of the diagenetic regime.

The hypothesis that the samples underwent rock-buffered diagenesis is further supported by comparisons of the clumped isotope temperature of different calcite textures. We measured the clumped isotope composition of the micrite adjecent to four brachiopods, and in all four cases, the temperature of the micrite was resolvably higher than the temperature of the brachiopod (Figure \ref{comparisons}). Micrite is expected to be less pristine than the brachiopod samples because micrite is fine-grained, not protected by a layer of brachiopod shell, and was not picked under a microscope prior to analysis. Because the less pristine material, micrite, was associated with higher temperatures, these comparisons are consistent with the hypothesis that the lowest temperature samples are the most pristine.

@@

The results of this study are also consistent with those of earlier clumped isotope studies. \cite{Came2007} and \cite{Finnegan2011} both measured the clumped isotope compositions of brachiopods from the Early Silurian on Anticosti Island, Canada. These two studies report temperatures of $30-35$\degrees C and seawater \deltao O$_{\text{SMOW}}$ values of around $-1\permil$, which are similar to the results of this study. Importantly, Anticosti Island recorded deposition on the eastern margin of the Panthalassic Ocean during the Silurian, while Gotland recorded deposition in the Rheic Ocean, which was separated from the Panthalassic Ocean by a mountain range. The strong similarity in temperatures and \deltao O between the two studies indicates that both studies likely reflect water that is representative of the global ocean, rather than an isolated inland sea.

Our results, along with those of previous clumped isotope studies \citep{Came2007,Dennis2013,Finnegan2011}, provide consistent explanations for the low \deltao O values of early Paleozoic carbonates. Carbonate clumped isotopes constrain the contribution of a change in seawater \deltao O to only a few permil of the observed trend, ruling out a value of $-8\permil$ for the \deltao O of early Paleozoic oceans \citep{Grossman2012}. Our results also serve as a reminder that many of the fossils preserved in the geologic record lived in tropical oceans. The \deltao O of carbonate in equilibrium with tropical ocean water should not be compared to the \deltao O of carbonate in equilibrium with water of average modern sea surface temperature, because tropical oceans are much warmer than the average sea surface temperature. Clumped isotopes additionally demonstrate that extremely hot ($\sim 70$\degrees C) ocean temperatures cannot be the cause of the depleted \deltao O values of early Paleozoic carbonates. Finally, clumped isotopes, as a highly sensitive indicator of diagenetic alteration, demonstrate that even among samples with original calcite fabric and pristine trace metal signatures, there is a large range in preservation quality. This means that the average fossil calcite in almost any sample suite is likely altered from its primary composition, and that diagenetic alteration is a key contributor to the low \deltao O values of early Paleozoic carbonates.

Due in large part to diagenetic alteration, our results are unable to resolve a change in temperature or the \deltao O of the oceans through the Ireviken event. The consistent scatter in fossil temperatures throughout the section likely rules out any large changes in temperature or \deltao O of the oceans, but this scatter prevents resolution of any smaller changes. Temperatures of 30\degrees C in tropical oceans are similar to those of modern tropical oceans, and therefore are not incompatible with the presence of polar ice sheets, as suggested by Early Silurian age tillites in Brazil \citep{Grahn1992}.

The results from this study join those of previous clumped isotope studies in implying that the \deltao O of seawater has been fairly constant throughout the Phanerozoic, at around $-1\permil$. Several studies have suggested that high-temperature alteration of the seafloor acts as a buffer for the \deltao O of seawater \citep{Muehlenbachs1976,Gregory1981,Gregory1991,Muehlenbachs1998}, but a more recent model implied that this buffer may not be strong enough to overcome slow changes in the \deltao O of seawater \citep{Jaffres2007}. If this buffer cannot explain the relative constancy of seawater \deltao O values through time, then the two competing controls on seawater \deltao O, alteration of rocks at high and low temperatures, must have remained in balance over the last 500 million years. This suggests that the magnitudes of high-temperature hydrothermal alteration and low-temperature weathering have not changed substantially, or that any changes in the magnitude of one process are tied through an as-yet unidentified feedback loop to balancing changes in the other.
